---
title: "Health and Economic consequences of weather events"
author: "Christoffer Haukvik"
date: "5 Dec 2017"
output: html_document
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path='figure/')

## Loading required libraries
library(dplyr)
library(ggplot2)
library(knitr)
library(reshape2)
```

## Health and Economic consequences of weather events

## Synopsis
This analysis uses weather data from x to evaluate the health- and economic consequences of weather in events in US from 1950 - 2011. The main findings are (a) bad weather is bad for health, (b) bad weather is bad for economics.

## Data processing
R has been used to collect and process the data in the following steps.

1. Data is is downloaded, extracted, and loaded in R data frame
```{r dataLoading, cache = TRUE}
if(!file.exists("data.csv.bz2")){
     fileURL <- 
     "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
     download.file(fileURL, dest="data.csv.bz2", 
                   mode="wb", method = "auto") 
}
data <- read.csv("data.csv.bz2")
```

2. Data transformation
In order to improve the processing of the data, we re-format dates
```{r dataPreProcessing}
# Formatting dates as dates
data$BGN_DATE <- as.Date(data$BGN_DATE, "%m/%d/%Y")
data$END_DATE <- as.Date(data$END_DATE, "%m/%d/%Y")

```


## Results
The analysis has addressed two specific questions:

### 1. Across the United States, which types of events are most harmful with respect to population health?

The data we are evaluating is measuring how weather events indirectly cause fatalities and injuries.

Population harm has been measured in two categories; number of fatalities, and number of injuries. Since the former is the more serious of the two, we have chosen to primarily investigate which weather events are most fatal.

To do this, we first summarize the measurements to display a table of the top 10 events which cause the most fatalities. For comparison, we also include the number of injuries which are associated to these events. In order to give an estimate of the percentage risk associated to any given event, we have also included statistics on the mean number of fatalities and injuries are associated with each event.

```{r healthStats}
# Subsetting events according to fatalities and injuries

healthData <- data %>%
     select(EVTYPE, FATALITIES, INJURIES) %>%
     group_by(EVTYPE) %>%
     summarize(Events = length(EVTYPE),
               Fatalities = sum(FATALITIES),
               Injuries = sum(INJURIES),
               MeanFatalities = round(mean(FATALITIES), 2),
               MeanInjuries = round(mean(INJURIES), 2)) %>%
     arrange(desc(Fatalities))

#1. Table of top 10 fatality and injury causes
kable(head(healthData, 10), row.names = TRUE, 
      col.names = c("Weather event", "Events", "Fatalities", "Injuries", 
                    "Fatalities (mean)", "Injuries (mean)"), 
     caption = "Weather events with greatest health effects")
```
As the table illustrates, tornados are by far the weather event which has the most fatal outcomes. Of the listed events, it also causes the most injuries. Nevertheless, on a per-event basis, it is not as fatal as other of the events listes (in particular heat related events).

```{r healthEvents}
#2. Creating facet grid with horizontal axis of most severe events for both fatalities and injuries

# Determining most severe 8 events in each category and combining them
topFatalities <- arrange(healthData, desc(Fatalities))$EVTYPE[1:8]
topInjuries <- arrange(healthData, desc(Injuries))$EVTYPE[1:8]
topEvents <- unique(combine(topFatalities, topInjuries))

# Subsetting health data based on these events
topSeverity <- subset(healthData, EVTYPE %in% topEvents,
                      select = c(EVTYPE, Fatalities, Injuries))

# Melting fatalities/injuries into factors
healthEvents <- 
     melt(topSeverity, id.vars = "EVTYPE", 
          measure.vars = c("Fatalities", "Injuries"), 
          value.name = "Incidents", variable.name = "Category")

# Ranking events based on fatalities, then injuries
healthEvents <- cbind(healthEvents, Rank = seq(1:nrow(healthEvents)))

# Presenting data in plot
ggplot(data = healthEvents, aes(x=reorder(EVTYPE, -Rank), 
                                y=Incidents, fill=EVTYPE)) +
     geom_col() + 
     geom_text(aes(label=Incidents), size = 3, position = "fill", 
               hjust = -1.4) +
     guides(fill=FALSE) +
     theme_light() +
     coord_flip() +
     facet_wrap(~Category, scales = "free_x") +
     theme(strip.background = element_rect(colour="black", fill="black")) +
     labs(title = "Severity of weather events on health in the US 1951-2011",
          x = "Weather event", y = "Number of Health incidents")
```

As we can see, tornados have the most sever effect on health effects, both in terms of fatalities and injuries. Second, excessive heat is another major factor in health effects.


### 2. Across the United States, which types of events have the greatest economic consequences?

```{r econStats}

# Extracting relevant data
econData <- subset(data, select = c(EVTYPE, PROPDMG, PROPDMGEXP, 
                                    CROPDMG, CROPDMGEXP), na.rm = TRUE)

# Cleaning relevant data

# Creating dataframe for converting known damange expense multipliers
convExp <- data.frame(expUnit = c("k", "K", "m", "M", "b", "B"),
                      Factor = c(10^3, 10^3, 10^6, 10^6, 10^9, 10^9))

# Collecting the EXP (cash multiplier) variables into one
expUnit <- c(as.character(econData$PROPDMGEXP), 
              as.character(econData$CROPDMGEXP))

# Melting dataframe
econDF <- melt(econData, id.vars = "EVTYPE",
               measure.vars = c("PROPDMG", "CROPDMG"),
               value.name = "Expense", variable.name = "DamageType")

# Assigning units to the melted dataframe
econDF <- cbind(econDF, expUnit)

# Excluding any 0 expenses
econDF <- econDF[econDF$Expense > 0, ]

# Creating DF which containins ONLY known expense units
econEvents <- inner_join(econDF, convExp, by="expUnit")

# Getting some info on the events which were excluded
exclEvents <- anti_join(econDF, convExp, by="expUnit")
exclNo <- nrow(exclEvents)

# Calculating expenses for each event
econEvents <- mutate(econEvents, fullExpense = Expense*Factor)

# Table of total expenses
totalExp <- econEvents %>%
     select(EVTYPE, fullExpense) %>%
     group_by(EVTYPE) %>%
     summarize(Events = length(EVTYPE),
               Expenses = sum(fullExpense),
               MeanExp = round(mean(fullExpense),0)) %>%
     arrange(desc(Expenses))

# Top 10 Economic effect weather events
top10Econ <- head(totalExp, 10)

kable(top10Econ, row.names = TRUE, 
      col.names = c("Weather event", "Events", "Total expenses (USD)", 
                    "Mean expense per event (USD"), 
     caption = "Weather events with greatest economic effects")

# Graph of total expenses

# Grouping top10 events for graph
econSummary <- econEvents %>%
     select(EVTYPE, DamageType, fullExpense) %>%
     filter(EVTYPE %in% top10Econ$EVTYPE) %>%
     group_by(EVTYPE, DamageType) %>%
     summarize(Events = length(EVTYPE),
               Expenses = sum(fullExpense),
               MeanExpense = mean(fullExpense)) %>%
     arrange(desc(Expenses))

ggplot(data = head(econSummary, 10), 
       aes(x=EVTYPE, y=Expenses, fill=DamageType)) +
     geom_col(position = "stack") +
     geom_text(aes(label = Expenses)) +
     theme_light() +
     labs(title = "Weather events with highest economic expense in the US", 
          x = "Weather event", y = "Total expenses")

```


## Summary


## References
Link to data file
Link to documentation
Link to website