---
title: "Tornados and floods are the weather events with the highest impact on US health and economy in the period 1950 to 2011"
author: "Christoffer Haukvik"
output: 
  html_document: 
    fig_caption: yes
keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.path='figure/')
```

## Synopsis
This data analysis has used the data published by the National Weather Service to determine which weather events have had the greatest health- and economic consequences in the US from 1950 until 2011.  
We seeks to answer two questions: (1) Across the United States, which types of events are most harmful with respect to population health, and (2) Across the United States, which types of events have the greatest economic consequences?  
To determine the greatest health impact of weather events, we have extracted the weather events and their associated fatalities and injuries for the entire period, and sorted the weather events' severity according to number of fatalities. To determine the economic impact of weather events, we have extracted the weather events and their associated expense costs in terms of property- and crop damage, and combined these to determine the total expense of each weather event.  
The results show that tornados by far cause the worst health effects (recorded 5633 fatalities and 91346 injuries), followed by excessive heat (recorded 1903 fatalities and 6525 injuries) and flash floods (978 fatalities and 1777 injuries). For economic effects, floods have the greatest total expenses (USD 150 billions), followed by hurricanes/typhoons (USD 719 billions) and tornados (USD 574 billions).  
Our recommendation is to ensure there are plans in place to minimise the effects weather events of tornados and floods have on health and economy.


## Data processing
In this section, we describe exactly how the data has been collected and processed into the presented results. We have used R (v3.4.2) to do this.

#### 1. Loading required libraries

This analysis will use the libraries below in manipulating- and plotting the data.
```{r packageLoading}
library(dplyr, warn.conflicts = FALSE)
library(ggplot2)
library(knitr)
library(reshape2)
```


#### 2. Data downloading, extraction, and loading

Data is downloaded from the given link and loaded into a data frame.
```{r dataLoading, cache = TRUE}
if(!file.exists("data.csv.bz2")){
     fileURL <- 
     "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
     download.file(fileURL, dest="data.csv.bz2", 
                   mode="wb", method = "auto") 
}
data <- read.csv("data.csv.bz2")
```


#### 3. Processing health-related data

We extract the health related data (fatalities and injuries) and the weather events (EVTYPE) from the initial dataset.  Second, we summarize the data per weather event, including the number of times the event has occurred, the totals for fatalities and injuries per event, and the mean number fo fatalities and injuries per event.

Finally, we order the data according to how many fatalities are associated with each weather event type (from highest to lowest number).

```{r healthProcessing}
# Subsetting events according to fatalities and injuries
healthData <- data %>%
     select(EVTYPE, FATALITIES, INJURIES) %>%
     group_by(EVTYPE) %>%
     summarize(Events = length(EVTYPE),
               Fatalities = sum(FATALITIES),
               Injuries = sum(INJURIES),
               MeanFatalities = round(mean(FATALITIES), 2),
               MeanInjuries = round(mean(INJURIES), 2)) %>%
     arrange(desc(Fatalities))
```

To get some further insight into the relationship between fatalities and injuries associated with weather events, we identify the top eight weather events with highest number of both fatalities and injuries. 

We then combine these weather events into a new dataset where we again rank according to number of fatalities suffered by the weather event.

```{r healthRanking}
# Determining most severe 8 events in each category and combining them
topFatalities <- arrange(healthData, desc(Fatalities))$EVTYPE[1:8]
topInjuries <- arrange(healthData, desc(Injuries))$EVTYPE[1:8]
topEvents <- unique(combine(topFatalities, topInjuries))

# Subsetting health data based on these events
topSeverity <- subset(healthData, EVTYPE %in% topEvents,
                      select = c(EVTYPE, Fatalities, Injuries))

# Melting fatalities/injuries into factors
healthEvents <- 
     melt(topSeverity, id.vars = "EVTYPE", 
          measure.vars = c("Fatalities", "Injuries"), 
          value.name = "Incidents", variable.name = "Category")

# Ranking events based on fatalities, then injuries
healthEvents <- cbind(healthEvents, Rank = seq(1:nrow(healthEvents)))
```


#### 4. Processing economics-realted data

We extract the economic data associated with each weather event; namely, the property- and crop expenses suffered by the event.
```{r econExtraction}
# Extracting relevant economic data
econData <- subset(data, select = c(EVTYPE, PROPDMG, PROPDMGEXP, 
                                    CROPDMG, CROPDMGEXP))
```

We also define and extract the identifiers for the unit of expense for each USD expense category (according to the National Weather Service; these should be "K" for thousand, "M" for millions, and "B" for billions). Since it is too uncertain to make a guess what the other labels might mean, we have excluded them from the analysis. 

We extract data into a new tidy dataset, where we use the above labels to multiply the expense (e.g. 150.00) with the factor (e.g. 1000 for "K") to later on being able to determine the actual dollar value (e.g. 150.000).
```{r factorProcessing}
# Creating dataframe for converting known damange expense multipliers
convExp <- data.frame(expUnit = c("k", "K", "m", "M", "b", "B"),
                      Factor = c(10^3, 10^3, 10^6, 10^6, 10^9, 10^9))

# Extracting the EXP (cash multiplier) variables into character vector
expUnit <- c(as.character(econData$PROPDMGEXP), 
              as.character(econData$CROPDMGEXP))

# Melting dataframe into tidy data with event types and expenses in rows
econDF <- melt(econData, id.vars = "EVTYPE",
               measure.vars = c("PROPDMG", "CROPDMG"),
               value.name = "Expense", variable.name = "DamageType")

# Assigning units to the melted dataframe
econDF <- cbind(econDF, expUnit)
```

Since we are not interested in any items where the expense is zero, we discard those from the analysis.

We then combine the expense factors (e.g. "K" for 1000) to the data and use this to calculate the actual dollar amount for each event.

```{r econProcessing, warning=FALSE}
# Excluding any 0 expenses
econDF <- econDF[econDF$Expense > 0, ]

# Creating DF which containins ONLY known expense units
econEvents <- inner_join(econDF, convExp, by="expUnit")

# Getting some info on the events which were excluded
exclEvents <- anti_join(econDF, convExp, by="expUnit")
exclNo <- nrow(exclEvents)

# Calculating expenses for each event
econEvents <- mutate(econEvents, fullExpense = Expense*Factor)
```

Finally, we group the data according to which weather events have the highest expenses, and include statistics on how many times each weather event has occurred, and what is the mean expense per event occurrence. 

Since we are most concerned with the weather events causing the highest economic impact, we create a final dataset including the 10 most expensive events.
```{r econSummarizing}
# Summarising data of total expenses
totalExp <- econEvents %>%
     select(EVTYPE, fullExpense) %>%
     group_by(EVTYPE) %>%
     summarize(Events = length(EVTYPE),
               Expenses = sum(fullExpense),
               MeanExp = round(mean(fullExpense),0)) %>%
     arrange(desc(Expenses))

# Top 10 Economic effect weather events
top10Econ <- head(totalExp, 10)
```

## Results
Having processed the data as above, we return to answer the two questions this analysis has been about.

#### 1. Across the United States, which types of events are most harmful with respect to population health?

Population harm has been measured in two categories; number of fatalities, and number of injuries. Since the former is the more serious of the two, we have chosen to primarily investigate which weather events are most fatal.

To do this, we first summarize the measurements to display a table (Figure 1) of the top 10 events which cause the most fatalities. For comparison, we also include the number of injuries which are associated to these events. In order to give an estimate of the risk associated to any given event, we have also included statistics on the mean number of fatalities and injuries are associated with each event.
```{r healthTable}
#Figure 1: Table of top 10 fatality and injury causes
kable(head(healthData, 10), row.names = TRUE, 
      col.names = c("Weather event", "Events", "Fatalities", "Injuries", 
                    "Fatalities (mean)", "Injuries (mean)"), 
     caption = "Figure 1: Weather events with greatest health effects")
```

As the table illustrates, tornados are by far the weather event with the most fatal outcomes. Of the listed events, it also causes the most injuries. Nevertheless, on a per-event basis, it is not as fatal as other of the events listes (in particular heat related events). However, it also ranks as the weather event with the greatest amount of injuries.

To illustrate this visually, please look at the the below graph (Figure 2):
```{r healthGraph}
# Presenting data in graph
ggplot(data = healthEvents, aes(x=reorder(EVTYPE, -Rank), 
                                y=Incidents, fill=EVTYPE)) +
     geom_col() + 
     geom_text(aes(label=Incidents), size = 3, position = "fill", 
               hjust = -1.4) +
     guides(fill=FALSE) +
     theme_light() +
     coord_flip() +
     facet_wrap(~Category, scales = "free_x") +
     theme(strip.background = element_rect(colour="black", fill="black")) +
     labs(title = "Figure 2: Severity of weather events on health",
          x = "Weather event", y = "Number of Health incidents")
```

As we can see, tornados have the most severe effect on health effects, both in terms of fatalities and injuries. Second, excessive heat is another major factor in health effects.

If the goal is to reduce the number of both fatalities and injuries as suffered by weather events, we would therefore recommend having a view on how to prevent fatalities and injuries suffered from tornado events.  

#### 2. Across the United States, which types of events have the greatest economic consequences?

Economic consequences have been categorized as both property- and crop expenses. The table below (Figure 3) summarizes which weather events have the greatest economic consequences, and also gives insight into how many of the weather events have been recorded, as well as the mean expense per occurrence.
```{r econResults}
kable(top10Econ, row.names = TRUE, 
      col.names = c("Weather event", "Events", "Total expenses (USD)", 
                    "Mean expense per event (USD)"), 
     caption = "Figure 3: Weather events with greatest economic effects")
```

As can be seen, floods have had the greatest economic impact for the given period, with more than twice the expense of other weather events. Nevertheless, they have a mean expense which is far below other events, such as hurricanes/tornados. Overall, we can also see that the main type of expense is related to property damages. 

If the goal is to minimize the economic impact of weather events, we would advice to have a plan for mitigating the effects of floods in general. However, given the significant effect of hurricanes/tornados when they first strike, one might be adviced to plan for these events as well if situated in a particularly exposed area.

## Recommendation

In this analysis we have investigated the impact weather events in the US have had on population health and economic expenses between 1950 and 2011. 

Tornados have had the greatest negative effect on population health (both fatalities and injuries) and floods have by far caused the overall greatest economic expense.

To minimize the impact of weather events on population health and economic expenses, we advice to have a preparedness plan for these events, including a contingency plan in place for when they occur.


## External links to National Weather Service
National Weather Service [documentation](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf)

National Weather service [FAQ](https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf)

Data used for analysis:  [Link](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2)